How crisis affect the low-income and high-income economies differently

### 2024-04-18 paper/paper.qmd
Fixing grammer

---
title: "How crisis affect the low-income and high-income economies differently"
subtitle: "An example from the 2020/2021 Covid-19"
author: 
  - Sean Liu 
thanks: "Code and data are available at: https://github.com/Sluuu/crisis-affect-countries.git"
date: today
date-format: long
abstract: "This paper examines the impacts of global crises focusing on the 2020 and 2021 COVID-19 pandemic's effects on low and high-income countries. It highlights how globalization links economies, leading to shared challenges during worldwide disturbances. The analysis reveals that low-income countries have suffered more acutely, experiencing higher unemployment, inflation, and slower GDP growth."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

In this paper, my focus is on the varying impacts of global crises on low and high income countries. I have selected the 2020 and 2021 COVID-19 pandemic as the primary subject of my study. This crises have led to varying degrees of economic uncertainty and fluctuations in economic performance worldwide.[@compare] The World Bank indicates that the COVID-19 pandemic has widen the inequality between low and high-income countries, with low-income countries experiencing higher unemployment and inflation rates, along with slower GDP growth. These trends align with the expectations set in my study.[@inequality]
```{r}
#| echo: false
#| warning: false
#| message: false

@@

Lastly, the data presented in @fig-inflation, @fig-gdp-growth, and @fig-ur, I observed that high-income countries shared similar economic performance trends. Despite not necessarily excelling in all economic indicators than the low-income countries, however, these countries generally maintain a more stable economic condition, with only a few exceptions or outliers noted. Which I will be expecting the data to be normally distributed. Also the overall economic performance in 2021 are better than in 2020.

@@

The model is set up by using R [@citeR] with the help of the following packages, `tidyverse` package of @tidyverse, `readr` package of @readr, `dplyr` package of @dplyr, `dataverse` package of @dataverse, `stargazer` and package of  @stargazer, and `broom` package of @broom.
Given that countries are classified as low or high income based on their GNI per capita, `gni_per_cap`, I will use this as the dependent variable in a multiple linear regression model to analyze the impact of various factors on `gni_per_cap` across these groups. I will explore how different variables affect `gni_per_cap` in both low and high-income countries. Specifically, I anticipate that low-income countries will experience larger negative effect from unemployment and inflation rates, as well as experience negative GDP growth. This analysis aims to provide insights into the economic challenges faced by these countries and inform potential policy interventions.

@@

The Labor force participation rate (`lf_par_rate`) coefficient is -52.818 with a standard error of 13.104, and it is significant at the p<0.01 level. This significant negative relationship implies that higher labor force participation is associated with lower GNI per capita among the low-income countries in this analysis, which might suggest underemployment or low-quality jobs prevalent in these economies.
The Trade percentage of GDP (`trade_per`) coefficient is 11.697 with a standard error of 3.992, significant at the p<0.01 level. This significant positive relationship suggests that a higher trade percentage of GDP is associated with higher GNI per capita, indicating the beneficial impact of trade on national income in low-income countries.

@@

The high income-countries have an overall better performance than the low-income countries: 
The GDP Growth (`gdp_growth`) coefficient is 567.121 with a standard error of 244.367, significant at the p<0.05 level. This suggests a positive and statistically significant relationship between GDP growth and GNI per capita in high-income countries, indicating that economic growth positively impacts income levels. Then the Unemployment Rate (`unem_rate`) coefficient is -849.519 with a standard error of 350.854, significant at the p<0.05 level. This indicates a significant negative impact of higher unemployment rates on GNI per capita, suggesting that unemployment is a critical economic issue affecting income in high-income countries. The Labor Force Participation Rate (`lf_par_rate`) coefficient is 433.735 with a standard error of 229.487, significant at the p<0.1 level. This shows a positive relationship between labor force participation and GNI per capita, though the relationship is less statistically significant. The Trade Percentage of GDP (`trade_per`) coefficient is 59.875 with a standard error of 22.554, significant at the p<0.01 level. This indicates a robust positive relationship, suggesting that trade openness contributes significantly to GNI per capita in high-income countries. Lastly, the Inflation (`inflation`) coefficient is -2,799.239 with a standard error of 600.123, significant at the p<0.01 level. This suggests a strong negative impact of inflation on GNI per capita, indicating that higher inflation rates are detrimental to income levels.

@@

## Model Adjustment
For a better performance and interpretation of the model, I have applied the log transformation to my model. The log transformation is to reduce the skewness in variables that have a skewed distribution, such as `gni_per_cpa`, `gdp_growth`, `population`, etc. By transforming these variables, the distribution is more symmetric, which is closer to the normal distribution. The Log transformations also help stabilize the variance across different levels of input variables, reduces homoscedasticity.
The log transformation variables are the follows:
Log of `gni_per_cap` The dependent variable's log transformation can help in normalizing the distribution and making relationships more linear. 
The log of `gdp_growth` and `population` shared similar reasons. As Economic growth rates and population can vary widely, and the impact on GNI per capita might scale logarithmically rather than linearly.
The following formula is the updated log transformation on certain variables.

@@

From the model we could see that in low-income countries the labour force participation rate has a significant negative value. Which in the high-income countries, the labour force participation rate has a significant positive value. This contrast may indicate that in low-income countries, increases in labor force participation may include low-quality, low-income jobs. Which are more unsecured during the COVID-19 crisis. Whereas in high-income countries, higher participation could be linked to more high-skill roles. [@job]
The Trade as Percentage of GDP has positive impact and significant in both low and high income countries. Although the trade decreases sharply in 2020, however, the recovery in 2021 was more significant than the dropped in 2020. Causing the  total trade flows are above the pre-pandemic levels.[@Trade]

The GDP Growth played an important role in the high-income countries and it has a significant and positive relationship. Possibly reflecting more developed financial systems and infrastructure which cause more growth into GNI per capita. However, it has insignificant negative relationship in the low-income coutries.

@@

The negative coefficient for GDP growth in low-income countries was expected however, it was not significant. For high-income countries, GDP growth shows a positive and significant impact, indicating robust economic dynamics where growth directly enhances national income levels. Also the high income countries showed a significant stronger positive relationship on labour force participation rate and trade as a share of GDP. 
From @fig-qq-low we could see that the observations were mostly separated and with multiple outliers which indicates that we have a low sample size also deviates from normality. Which shows that for the low-income countries model, it has potentially leading to skewed or biased estimates.
From @fig-qq-high the points on the Q-Q plot fall approximately along a straight line, suggesting that the residuals follow a normal distribution. Which it also has less outliers compared with the low-income model. It also furthe confirm that the residuals are normally distributed, supporting the validity of the least squares regression assumptions.

@@

There are several weaknesses of this study, which in the low-income countries model several important variables like GDP growth, unemployment, and inflation did not show statistically significant effects. This might be due to the lack of sample sizes and the high fluctuations in the data. The second weakness was the lack of variables, the model did not include some relevant variables that affect GNI per capita, such as political stability, technology, and education levels, etc. Which both results might create biases onto the results.
The next step of the study might be on increase the data size, such as include more years or some micro-level data.

>>>>

---
title: "How crisis affect the low-income and high-income economies differently"
subtitle: "An example from the 2020/2021 Covid-19"
author: 
  - Sean Liu 
thanks: "Code and data are available at: https://github.com/Sluuu/crisis-affect-countries.git"
date: today
date-format: long
abstract: "This paper examines the impacts of global crises focusing on the 2020 and 2021 COVID-19 pandemic's effects on low and high-income countries. It highlights how globalization links economies, leading to shared challenges during the worldwide pandemic. The analysis shows that low-income countries have suffered more than high-income countries, by experiencing higher unemployment, inflation, and slower GDP growth."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

In this paper, my focus is on the different impacts of global crises on low and high income countries. I have selected the 2020 and 2021 COVID-19 pandemic as the primary subject of my study. This crises have led to varying degrees of economic uncertainty and fluctuations in economic performance worldwide.[@compare] The World Bank indicates that the COVID-19 pandemic has widen the inequality between low and high-income countries, with low-income countries experiencing higher unemployment and inflation rates, also with slower GDP growth. These trends align with the expectations set in my study.[@inequality]
```{r}
#| echo: false
#| warning: false
#| message: false

@@

Lastly, the data presented in @fig-inflation, @fig-gdp-growth, and @fig-ur, I observed that high-income countries shared similar economic performance trends. Although not necessarily always better in all economic indicators than the low-income countries, however, these countries generally maintain a more stable economic condition, with only a few exceptions or outliers noted. Therefore, I will be expecting the data to be normally distributed. Also the overall economic performance in 2021 are better than in 2020.

@@

The model is set up by using R [@citeR] with the help of the following packages, `tidyverse` package of @tidyverse, `readr` package of @readr, `dplyr` package of @dplyr, `dataverse` package of @dataverse, `stargazer` and package of  @stargazer, and `broom` package of @broom.
Given that countries are classified as low or high income based on their GNI per capita, `gni_per_cap`, I will use this as the dependent variable in a multiple linear regression model to analyze the impact of various factors on `gni_per_cap` across these groups. I will explore how different variables affect `gni_per_cap` in both low and high-income countries. Specifically, I expect that low-income countries will experience larger negative effect from unemployment and inflation rates, as well as experience negative or slower GDP growth. This analysis aims to provide insights into the economic challenges faced by these countries and inform potential policy impacts.

@@

The Labor force participation rate (`lf_par_rate`) coefficient is -52.818 with a standard error of 13.104, and it is significant at the p<0.01 level. This significant negative relationship shows that higher labor force participation is connected with lower GNI per capita among the low-income countries in this analysis, which might suggest underemployment or low-quality jobs in these economies.
The Trade percentage of GDP (`trade_per`) coefficient is 11.697 with a standard error of 3.992, significant at the p<0.01 level. This significant positive relationship suggests that a higher trade percentage of GDP is associated with higher GNI per capita, indicating the beneficial impact of trade on national income in low-income countries.

@@

The high income-countries have an overall better performance than the low-income countries: 
The GDP Growth (`gdp_growth`) coefficient is 567.121 with a standard error of 244.367, significant at the p<0.05 level. This suggests a positive and statistically significant relationship between GDP growth and GNI per capita in high-income countries, indicating that economic growth positively impacts income levels. Then the Unemployment Rate (`unem_rate`) coefficient is -849.519 with a standard error of 350.854, significant at the p<0.05 level. This indicates a significant negative impact of higher unemployment rates on GNI per capita, suggesting that unemployment is a critical economic issue affecting income in high-income countries. The Labor Force Participation Rate (`lf_par_rate`) coefficient is 433.735 with a standard error of 229.487, significant at the p<0.1 level. This shows a positive relationship between labor force participation and GNI per capita, though the relationship is less statistically significant. The Trade Percentage of GDP (`trade_per`) coefficient is 59.875 with a standard error of 22.554, significant at the p<0.01 level. This shows a robust positive relationship, suggesting that trading contributes significantly to GNI per capita in high-income countries. Lastly, the Inflation (`inflation`) coefficient is -2,799.239 with a standard error of 600.123, significant at the p<0.01 level. This suggests a strong negative impact of inflation on GNI per capita, indicating that higher inflation rates are fluctuating income levels.

@@

## Model Adjustment
For a better performance and interpretation of the model, I have applied the log transformation to my model. The log transformation is to reduce the skewness in variables that have a skewed distribution, such as `gni_per_cpa`, `gdp_growth`, `population`, etc. By transforming these variables, the distribution is more symmetric, which is closer to the normal distribution. The Log transformations also help stabilize the variance across different levels of input variables, reduces homoscedasticity.
The log transformation variables are the follows:
Log of `gni_per_cap` The dependent variable's log transformation can help in normalizing the distribution and making relationships more linear. 
The log of `gdp_growth` and `population` shared similar reasons. As economic growth rates and population can vary widely, and the impact on GNI per capita might scale logarithmically other than linearly.
The following formula is the updated log transformation on certain variables.

@@

From the model we could see that in low-income countries the labour force participation rate has a significant negative value. Which in the high-income countries, the labour force participation rate has a significant positive value. This contrast may indicate that in low-income countries, increases in labor force participation may include low-quality, low-income jobs. Which are more unsecured during the COVID-19 crisis. Whereas in high-income countries, higher participation could be linked to more high-skilled roles. [@job]
The Trade as Percentage of GDP has positive impact and significant in both low and high income countries. Although the trade decreases sharply in 2020, however, the recovery in 2021 was more significant than the dropped in 2020. Causing the total trade flows are above the pre-pandemic levels.[@Trade]

The GDP Growth played an important role in the high-income countries and it has a significant and positive relationship. Possibly reflecting more developed financial systems and infrastructure which cause more growth into GNI per capita. However, it has insignificant negative relationship in the low-income countries.

@@

The negative coefficient for GDP growth in low-income countries was expected however, it was not significant. For high-income countries, GDP growth shows a positive and significant impact, indicating strong economic dynamics and growth directly improving national income levels. Also the high income countries showed a significant stronger positive relationship on labour force participation rate and trade as a share of GDP. 
From @fig-qq-low we could see that the observations were mostly separated and with multiple outliers which indicates that we have a low sample size also deviates from normality. Which shows that for the low-income countries model, it has a probability of leading to skewed or biased estimates.
From @fig-qq-high the points on the Q-Q plot fall approximately along a straight line, suggesting that the residuals follow a normal distribution. Which it also has less outliers compared with the low-income model. It also further confirm that the residuals are normally distributed.

@@

There are several weaknesses of this study, which in the low-income countries model several important variables like GDP growth, unemployment, and inflation did not show statistically significant effects. This might be due to the lack of sample sizes and the high variations in the data. The second weakness was the lack of variables, the model did not include some relevant variables that affect GNI per capita, such as political stability, technology, and education levels, etc. Which both results might create biases onto the results.
The next step of the study might be on increase the data size, such as include more years or some micro-level data.
